The Effects of Varying Depth in Cosmic Shear Surveys

### 2019-05-06 appendix.tex
First semi-complete version after recent changes. Will still be subject to many modifications.

>>>>

\subsection{The function $E(\theta)$}
\label{sec:model_e}
When computing the shear correlation between a pair of galaxies, it is of central importance whether those two galaxies lie in the same pointing or not. We want to model the probability that a pair of galaxies with separation $\b\theta$ lie in the same pointing by the function $E(\b\theta)$, which we will derive here:

{
\def\vec{\b}
\begin{figure}
    \centering
    \def\svgwidth{200pt}    
    \hspace*{1cm}
    \input{images/eoftheta_new.pdf_tex}  
    \hspace*{-2cm}
    \caption[Graphic how to obtain $E(\b\theta)$]{Graphic representation on how to obtain the function $E(\b\theta)$. For a separation vector $\b\theta$, the dashed square represents the area of galaxies that have their partner in the same pointing.}
    \label{fig:explain_etheta}
\end{figure}
}
Given one square field of length $L$ (in our case $L=60\arcmin$) and a separation vector $\b\theta$, without loss of generality we can assume $\theta_1,\theta_2\geq 0$. As depicted in Figure \ref{fig:explain_etheta}, the dashed square represents all possible positions that the first galaxy can take, such that the second galaxy is still within the same pointing. The volume of this square equals \begin{equation}
V(|\b\theta|,\phi)  = \big[L-|\b\theta|\cos(\phi)\big]\,\big[L-|\b\theta|\sin(\phi)\big]\, ,
\end{equation} where $\phi$ represents the angle of the vector $\b\theta$. The function $E(\b\theta)$ then simply equals $V(|\b\theta|,\phi)/L^2$. To exclude negative Volumes (which could occur when $|\b\theta|>1$ holds), we need to add the Heaviside theta function $\mathcal{H}$:
\begin{equation}
E(\b\theta)  = \left[1-\frac{|\b\theta|}{L}\cos(\phi)\right]\,\left[1-\frac{|\b\theta|}{L}\sin(\phi)\right]\, \mathcal{H}\left[1-\frac{|\b\theta|}{L}\cos(\phi)\right]\,\mathcal{H}\left[1-\frac{|\b\theta|}{L}\sin(\phi)\right]\, .
\label{eq:eoftheta1}
\end{equation} 
As $E(\b\theta)$ is not isotropic, in order to obtain the function $E(\theta) = E(|\b\theta|)$, we need to azimuthally average Equation \eqref{eq:eoftheta1} over all angles $\phi$. While the case $\theta_1,\theta_2\geq 0$ certainly does not hold for all angles $\phi$, we can eliminate the other cases by simple symmetry.
\begin{align}
E(\theta) = & \frac{4}{2\pi}\int_0^{\frac{\pi}{2}}\d\phi\, E(\b\theta) = \frac{2}{\pi}\begin{cases}
\int_0^{\frac{\pi}{2}} \d\phi\, \left[1-\frac{|\b\theta|}{L}\cos(\phi)\right]\,\left[1-\frac{|\b\theta|}{L}\sin(\phi)\right]\, ,  & |\b\theta|  \leq L \\[10pt]
\int_{\cos\inv(L/|\b\theta|)}^{\sin\inv(L/|\b\theta|)} \d\phi\, \left[1-\frac{|\b\theta|}{L}\cos(\phi)\right]\,\left[1-\frac{|\b\theta|}{L}\sin(\phi)\right] \, , \quad   & L \leq |\b\theta| \leq \sqrt{2}L \\[10pt]
0\, , &\sqrt{2}L \leq\theta
\end{cases} \nonumber\\[10pt]
 = & \begin{cases}
\frac{1}{L^2 \pi}\left[L^2\pi - (4L-\theta) \theta\right]\, ,  & \theta \leq L \\[10pt]
\frac{2}{\pi}\,\left[4\sqrt{\frac{\theta^2}{L^2}-1} -1 - \frac{\theta^2}{2L^2} - \cos\inv\left(\frac{L}{\theta}\right) + \sin\inv\left(\frac{L}{\theta}\right)\right]\, ,  & L  \leq \theta \leq \sqrt{2}L \\[10pt]
0\, ,  & \sqrt{2}L \leq \theta
\end{cases}\, .
\end{align}

### 2019-05-06 discussion.tex
First semi-complete version after recent changes. Will still be subject to many modifications.

\section{Discussion}
\label{sec:discussion}
With our semi-analytic model we try to describe the impacts of varying depth in ground-based surveys. During our analysis we have assumed a few simplifications, which we will discuss.
\begin{enumerate}
\item We implemented a variation in depth that only depends on the pointing. However, in reality galactic extinction, CCD imperfections, dithering strategies and other factors also influence the depth on different scales. Although the variation between pointings is the dominating factor, a complete analysis implementing those factors could yield a slightly different result.
\item In our main analysis we assumed an infinite number of fields. This ignores boundary effects that would arise in the vincinity of the edge of a footprint (there finding a galaxy outside of the same pointing would be less likely). Also, another effect is ignored: If one has a number of $N$ fields in each percentile, and one galaxy is in a pointing of a certain percentile, then the probability that a different pointing is of the \emph{same} percentile is reduced by the factor $(N-1)/N$. However, we modeled the second effect, and already at 100 pointings it barely makes any difference. We did not model the boundary effects, but we assume that they are likewise neglible. \todo{This needs to be rewritten to account for the new insights.}
\item We have assumed that the depth of neighbouring pointings is uncorrelated. While there is no reason not to assume that, we have not yet verified that this is indeed the case.
\item In our Monte Carlo simulations we have only had $\Omega_{\rm m}$ and $\sigma_8$ as free parameters. While we do not expect most other parameters to make a difference, baryonic feedback and/or intrinsic alignments also change the correlation functions on small scales, so it is definitely possible that in a Monte Carlo chain including free parameters for these effects, those would change to mitigate the varying depth, and the impact on the actual cosmological parameters would be much smaller.
\end{enumerate}
We have determined that the effects of varying depth are not significant for the KiDSV-450 survey. For a Euclid-like survey they will play an important role, but before a model is implemented in such a precise survey, one should revise the used simplifications, especially the first one. We suspect that it should be possible to account for different aspects of varying optical depth by just modifying the function $E(\theta)$.

However, for the parameter $S_8$ this effect is insignificant, which in particular means that a variation in depth can not explain the discrepancies between the analyses of the CMB and those of the local universe. It can also be responsible for the occurence of $B$-Modes, although only to an insignificant amount.

>>>>

\section{Discussion}
\label{sec:discussion}
With our semi-analytic model we try to describe the impacts of varying depth in ground-based surveys. During our analysis we have assumed a few simplifications, which we will discuss.
\begin{enumerate}
\item In the most general terms, we are analysing the effects of a position-based selection function on cosmic shear surveys. In our analysis, this selection function was governed by the $r$-band depth of a pointing. This neglects a number of other effects: The depth in different bands and the seeing of a pointing will also modify the number densities and redshift distributions on the scale of a pointing, whereas dithering strategies as well as imperfections in the telescope and CCD cause modifications on sub-pointing scales. However, we believe that these effects are subdominant compared to the variations caused by the $r$-band depth.

\item We have assumed an infinitely large survey area with an uncorrelated distribution of the depth-function. While the boundary effects arising from a finite survey footprint would have a small impact on the shape of the function $E(\theta)$\footnote{This would be due to the fact that a pointing next to a boundary has less neighbours, therefore making it more likely that a galaxy pair is in the same pointing.}, the governing factor is the shot noise of the depth-distribution. We have assumed that the probability that a neighbouring pointing is of percentile $n$ is exactly the expectation value, namely $1/10$. While this is true for an infinitely large survey with an uncorrelated distribution of the depth-function, Figure \ref{fig:all_xis} clearly shows that it is not valid for a $100\,\rm{deg}^2$ field. Whether our assumption holds true for the $450\,\rm{deg}^2$ footprint of the KV450 survey, or even the $1350\,\rm{deg}^2$ footprint of the final survey is an important question. Also, we have assumed an uncorrelated distribution of the depth-function, both in the simulations as well as in the models. While this is likely true as a rough approximation (very few pointings of the survey were taken on the same night and thus under the same weather conditions), effects like airmass, lunar phase, Galactic extinction and seasonal weather are likely to influence the depth on scales larger than one pointing.

An outlook for the mitigation of finite field effects, boundaries and a correlated distribution of depth is given in Appendix \ref{sec:expand_eoftheta}. As a preliminary result we find that finite field effects are not significant for a $450\,\rm{deg}^2$ or $1000\,\rm{deg}^2$-field, if the distribution of depth is uncorrelated.

\item In our MCMC simulations we did not account for degeneracies with other cosmological parameters
% (apart from $\Omega_{\rm m}, \sigma_8, w_0$ and $w_a$)
 or observational effects. Especially intrinsic alignments and baryonic feedback also modify the correlation functions especially on small scales, so they are probably degenerate with the effect of varying depth. In an actual MCMC simulation that accounts for these effects, we suspect that the parameters for intrinsic alignments and baryonic feedback change to mitigate this effect, and the impact on cosmological parameters is actually smaller than in our results. Also, possible degeneracies between $S_8$ and other cosmological parameters might bias the resulting values. %Furthermore, when including the effects of dark energy we notice that our observation of $S_8$ being constant is no longer valid: For a `Euclid-like' survey we notice a bias of $\sim 0.8\sigma$. This gives rise to the assumption that the results of an MCMC with the full parameter space, including observational effects, will differ from our results. 
\end{enumerate}

Despite these repercussions, we are confident to say that the effects of varying depth are not significant for the KV450 survey. The cosmological parameters did not change significantly and the main parameter, $S_8$, is especially robust against this effect. In particular this means that a varying depth can not explain the discrepancy between observations of the local Universe and results from analysis of the CMB.

We have shown that this effect can create B-modes. However, \citet{2018arXiv181002353A} measured the B-modes of the KV450 survey in the same $\theta$-range. Those B-modes are at least one order of magnitude larger and still consistent with zero, so it is safe to say that the modes created by varying depth are negligible. An interesting observation is that the change in E-modes is as big as the created B-modes (compare Figure \ref{fig:cosebis_eandb}). This means that as soon as this effect causes significant biases in the cosmological parameters, it will also create significant B-modes\footnote{While this is no big surprise, it is not trivial. It could be possible that a systematic effect only creates E-modes and no B-modes, which would be extremely unfortunate as it could bias cosmological parameters without ever being detected by a B-mode analysis.}. Additionally, the created pattern is very characteristic, which makes it easy to recognize in a B-mode analysis of an actual survey.

For next-generation surveys like Euclid, this effect will be significant. Although Euclid is a space-based telescope, the photometric redshift determination will still be done by ground-based telescopes and therefore suffer from the same effects. While we did not yet perform a quantitative study of this effect for Euclid, we are certain that a such a study should be conducted.

As an outlook it would be interesting to perform a full MCMC simulation, including both cosmological and nuisance parameters. One might also check the most popular extensions to the standard model of cosmology, and see whether varying depth could cause a bias towards one of these models. Furthermore, the method developed in Appendix \ref{sec:expand_eoftheta} could be applied to the actual footprint of the KV450 survey to see whether finite field effects or a correlated distribution of $r$-band depth are significant.
Additionally is interesting to note that $E(\theta)$ is the azimuthal average of the function $E(\b\theta)$, which is not isotropic. Therefore, it would be possible to observe a direction-dependent correlation function $\xi_\pm^{ij,\rm{obs}}(\b\theta)$ in future surveys. An anisotropy in the observed correlation function could be a sign for the influence of varying depth.

### 2019-05-06 introduction.tex
First semi-complete version after recent changes. Will still be subject to many modifications.

\section{Introduction}
The discovery of cosmic shear has provided us with a new and powerful cosmological tool to investigate the $\Lambda$CDM Model and determine its parameters. Contrary to the analysis of the Cosmic Microwave Background \citep{2018arXiv180706209P}, cosmic shear is more sensitive to the properties of the local Universe and thus provides an excellent consistency check for the standard model of cosmology. Current Cosmic Shear surveys are especially sensitive to the paramter $S_8=\sigma_8 \sqrt{\Omega_{\rm m}/0.3}$, where $\sigma_8$ denotes the normalisation of the matter power spectrum and $\Omega_{\rm m}$ is the matter density.
It is interesting to note that all three current major cosmic shear results report a lower $S_8$ than inferred from CMB analysis: While \citet{2018arXiv180706209P} determined a value of $S_8 = 0.830 \pm 0.013$, 
%\citet{2013MNRAS.432.2433H} report $S_8 = 0.759 \pm 0.020$ from analysis of CFHTLens data,
\citet{2018arXiv180909148H} report $S_8 = 0.800^{+0.029}_{-0.028}$ from analysis of the Subaru Hyper Suprime-Cam survey, \citet{2017MNRAS.465.1454H} report $S_8 = 0.745\pm 0.038$ from KiDS data and the Dark Energy Survey \citep{2018PhRvD..98d3528T} reports $S_8=0.782\pm 0.027$. Also, \citet{2013MNRAS.432.2433H} report $S_8 = 0.759 \pm 0.020$ from analysis of CFHTLens data. This discrepancy could be a statistical coincidence, a sign of new physics or the manifestation of an unknown systematic effect. \todo{Citations.. \citet{2017MNRAS.464.1676A} has a few...}

To check for remaining systematics, a weak lensing signal can be divided into two components, the so-called E- and B-modes \citet{2002ApJ...568...20C,2002A&A...389..729S}. To leading order, B-modes can not be created by astrophysical phenomena and are thus an excellent test for remaining systematics. As \citet{2017MNRAS.465.1454H} report the significant detection of B-modes, it is well motivated to check for possible systematics that are yet unaccounted for. Note that the non-existence of B-modes does not imply that the sample is free of remaining systematics.

One systematic effect is the variation of depth in a survey. While effects like Galactic extinction or dithering strategies do play a role in every survey, this work focuses on the effects caused by varying atmospheric conditions, that are found in ground-based surveys. To first order, this variation can be modelled by a step-like depth function that varies from pointing to pointing. In this work we assume the specifications of the Kilo-Degree Survey, namely a collection of $1\,\text{deg}^2$ square fields. Furthermore we neglect boundary effects.

>>>>

\section{Introduction}
The discovery of cosmic shear has provided us with a new and powerful cosmological tool to investigate the $\Lambda$CDM Model and determine its parameters. Contrary to the analysis of the CMB by \citet{2018arXiv180706209P}, cosmic shear is more sensitive to the properties of the local Universe and thus provides an excellent consistency check for the standard model of cosmology. Current cosmic shear surveys are especially sensitive to the parameter $S_8=\sigma_8 \sqrt{\Omega_{\rm m}/0.3}$, where $\sigma_8$ denotes the normalisation of the matter power spectrum and $\Omega_{\rm m}$ is the matter density.
It is interesting to note that all three current major cosmic shear results report a lower $S_8$ than inferred from CMB analysis: While \citet{2018arXiv180706209P} determined a value of $S_8 = 0.830 \pm 0.013$, 
%\citet{2013MNRAS.432.2433H} report $S_8 = 0.759 \pm 0.020$ from analysis of CFHTLens data,
\citet{2018arXiv180909148H} report $S_8 = 0.800^{+0.029}_{-0.028}$ from analysis of the Subaru Hyper Suprime-Cam survey, \citet{2018arXiv181206076H} report $0.737_{-0.036}^{+0.040}$ from KiDS+VIKING data and the Dark Energy Survey \citep{2018PhRvD..98d3528T} reports $S_8=0.782\pm 0.027$. Also, \citet{2013MNRAS.432.2433H} report $S_8 = 0.759 \pm 0.020$ from analysis of CFHTLens data. This discrepancy has received a lot of attention \citep{2013PDU.....2..166V}. It could be interpreted as a statistical coincidence, a sign of new physics like massive neutrinos \citep{2014PhRvL.112e1303B}, time-varying dark energy or modified gravity \citep{2016A&A...594A..14P}; or as the manifestation of a systematic effect, either in the cosmic shear surveys or in the Planck mission \citep{2016ApJ...818..132A}, that is not yet accounted for.

As weak gravitational lensing measures a tiny signal over a large sample, it is extremely sensitive to anything that systematically biases the measurements, such that the error bars in current surveys arise to equal parts from statistical and systematic uncertainties \citep[compare][]{2017MNRAS.465.1454H}. With next-generation surveys like the Large Synoptic Survey Telescope and Euclid right at the doorstep, systematic effects in gravitational lensing have received an unprecedented amount of attention \citep{2018arXiv181002353A,2019arXiv190207439B,2019arXiv190109488S}. %\todo{Citations.. \citet{2017MNRAS.464.1676A} has a few...}

To check for remaining systematics, a weak lensing signal can be divided into two components, the so-called E- and B-modes \citep{2002ApJ...568...20C,2002A&A...389..729S}. To leading order, B-modes can not be created by astrophysical phenomena and are thus an excellent test for remaining systematics. As \citet{2017MNRAS.465.1454H} reported the significant detection of B-modes, it was well motivated to check for possible systematics that are not yet accounted for. Note that the non-existence of B-modes does not necessarily imply that the sample is free of remaining systematics.

One systematic effect is the variation of depth in a survey. While effects like Galactic extinction or dithering strategies do play a role in every survey, this work focuses on the effects caused by varying atmospheric conditions, that are found in ground-based surveys. To first order, this variation can be modelled by a step-like depth function which varies from pointing to pointing. In this work we assume the specifications of the Kilo-Degree Survey, namely a collection of $1\,\text{deg}^2$ square fields.
